Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging

### 2018-10-11 HRUIonGPU/document.tex
Testing change of frame letters

In the next section we present background information on Capon beamforming in both element-space and beamspace. We also give an introduction to the GPU computing model. In Section \ref{sec:meth} we give a summary of our parallel implementation of the element-space Capon beamformer, before the GPU implementation of the beamspace Capon beamformer is presented in Section \ref{sec:bs}. Benchmark results and resulting images for both implementations are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}. Finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.

\section{Background}\label{background}

\subsection{Capon Beamforming}\label{sec:es-capon}

@@

The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS). Here, the signal recorded at element number $m \in [0,M-1]$ at time $n$ is delayed with an appropriate delay $\Delta_m[n]$ to focus and steer the ultrasound beam. This delayed data is here denoted $x_m[n]$. The beamformer output $z[n]$ is then the sum of all $M$ elements weighted with a factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n] = \vec{w}^H\vec{x}[n], \label{eq:z}
\end{align}
where $(\cdot)^H$ is the Hermitian transpose. The weight vector, or window, $\vec{w}$ is usually real in DAS beamforming, and is applied to trade resolution for a lower side lobe level. Although different windows can be used in different areas of the images, $\vec{w}$ is usually predefined.

As shown in Fig.\ \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.   The Capon beamformer produces weights that minimize the beamformer output power while maintaining unity gain in the focus direction. The result is that interference impinging from other directions is suppressed \cite{Synnevag2007}. As with coarse-fine beamformers \cite{Thomenius}, the complex weights cause a micro-steering which is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.

%Formally stated, the adaptive weights are found by solving the following minimization problem:
%\begin{align}
%&\min_{\vec{w}} E\{|z[n]|^2\} = \min_{\vec{w}}\vec{w}^H\mat{R}\vec{w}\label{eq:minz}\\
%&\text{subjected to } \vec{w}^H\vec{a} = 1,\label{eq:constraint}
%\end{align}
%where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$ is the spatial covariance matrix, and $\vec{a} = \left[ \begin{array}{cccc} e^{-jk_{\tilde{x}}\tilde{x}_0}& e^{-jk_{\tilde{x}}\tilde{x}_1}& ...& e^{-jk_{\tilde{x}}\tilde{x}_{M-1}}\end{array}\right] ^T$ is a far-field steering vector towards angle $\theta = \arcsin(\frac{\lambda k_{\tilde{x}}}{2\pi})$.
%The constraint in (\ref{eq:constraint}) is there to avoid the trivial solution $\vec{w} = 0$, and a distorted respons in the direction of $\vec{a}$. 
%The solution to the minimization problem in (\ref{eq:minz}) and (\ref{eq:constraint}) is

\begin{figure}
\centerline{\includegraphics[width=.7\linewidth]{gfx/beamforming_mv.eps}}
\caption{Capon beamforming. The impinging signal is demodulated, and aligned in phase by applying steering and focusing delays $\Delta_m[n]$. Adaptive weights, $\vec{w}[n]$, are then calculated based on in-phase data, and finally the output is formed by a Capon-weighted coherent sum.}
\label{fig:mvbf}
\end{figure}

The Capon beamformer, following the approach in \cite{Synnevag2009}, can be divided into three main steps: estimation of a sample covariance matrix, solving a system of linear equations, and calculation of the beamformer output.  The sample covariance matrix is estimated as 
%
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
\end{align}
%
where  $N_K = 2K + 1$ is the number of temporal samples included, $N_L = M-L+1$ is the number of subarrays, and $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l\text{th}$ subarray of length $L$. Note that the covariance matrix dimension is $L \times L$. The covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness:
%
\begin{align}\label{eq:diag}
\mat{\hat{R}}[n] = \mat{\breve{R}}[n] + \epsilon[n]\mat{I} = f(\vec{x}[n]).
\end{align}
%
The operator $f(\cdot)$ represents the process of estimating $\mat{\hat{R}}$ from the input $\vec{x}$.
A factor proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
%
\begin{align}\label{eq:diag_adapt}
\epsilon[n] &= d \times \frac{\text{trace}\{\mat{\breve{R}}[n]\}}{L},
\end{align}
%
has been applied in much of the recent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}. It has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.

Next, a linear system of equations has to be solved,
%
\begin{align}\label{eq:b}
\vec{b}[n] = \mat{\hat{R}}[n]^{-1}\vec{a} \in \mathbb{C}^L,
\end{align}
%
where the steering vector $\vec{a} = \vec{1}_L = [1_0, 1_1, ..., 1_{L-1}]^T$ since $\vec{x}$ is pre-delayed.

Finally the Capon weight vector is formed from (\ref{eq:b}) as
%
\begin{align}\label{eq:w}
\vec{w}[n] = \frac{\vec{b}[n]}{\vec{1}^H\vec{b}[n]} \in \mathbb{C}^L,
\end{align}
%
and the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights. This is formally stated as
%
\begin{align}
z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l[n] \label{eq:z_mv}\\
&= 1/N_L \, (\vec{w}[n] * \vec{1}_{N_L})^H\vec{x}[n] \label{eq:z_mv2},
\end{align}
%where (this matrix is wrong)
%\begin{align}
%\mat{C} = \left(
%\begin{matrix}
%\vec{1}_1 & \vec{1}_2 & \cdots & \vec{1}_{N_L} & \vec{0}_1 & \cdots & \vec{0}_{N_L-1} & \vec{0}_{N_L}\\
%\vec{0}_{N_L-1} & \vec{0}_{N_L-2} & \cdots & \vec{0}_{0} & \vec{1}_{N_L-1} & \cdots & \vec{1}_{2} & \vec{1}_{1}\\
%\end{matrix}
%\right)^T
%\end{align}
where $*$ is a non-truncated convolution across elements.
Note how we can either choose to sum all subarrays (\ref{eq:z_mv}), or accumulate a weighting per element (\ref{eq:z_mv2}).

In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krima}. Since medical ultrasound imaging is an active, short-pulsed system, there is a limited amount of data from which $\mat{\hat{R}}$ can be estimated. The estimated matrix is therefore often found to be both inaccurate and poorly conditioned. Another challenge with active systems is the high correlation between signal and interference, resulting in signal cancellation \cite{Reddy1987}. Thus, the beamformer output can be lower than expected, even when there is signal present and the distortionless-response criterion is applied.

@@

%\begin{figure}
%\includegraphics[width=3in]{gfx/beamforming_mv_lowres.png}
%\caption{}
%\label{fig:mv}
%\end{figure}

@@

\subsection{Beamspace Processing}\label{bs-capon}
Nilsen and Hafizovic \cite{Nilsen2009} have proposed to apply beamspace processing to reduce the required computations of the Capon beamformer when applied to medical ultrasound imaging. Beamspace is an alternative representation of the per-element data, transformed into a fan of beams covering the sector illuminated by transmit beams. The transformation is formally stated as
\begin{align}
\vec{x}_\text{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \ \ \ \ \ \ \ \ \ \ \, \mat{B} \in \mathbb{C}^{N,M} \label{eq:butler} \\ 
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M} \ \ \ N \le M.
\end{align} 
The steering vectors in $\mat{B}$, the so-called Butler or normalized discrete fourier transform (DFT) matrix, determine each beam's direction in space. Beams where no interference is present can therefore be removed by simply removing rows from $\mat{B}$.

@@

The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix 
\begin{align}
\mat{\hat{R}}_\text{BS} = f(\vec{x}_\text{BS}) = f(\mat{B}\vec{x}), 
\end{align}
where $\mat{B} \in \mathbb{C}^{N_b,L}$, $N_b$ is the number of selected beams, and $f(\cdot)$ represent the covariance estimation process as defined in (\ref{eq:diag}).

In the remaining sections, the methods described in this section and Section \ref{sec:es-capon} will be referred to as beamspace Capon (BS-Capon) and element-space Capon (ES-Capon) respectively. The name Capon will refer to both methods.

@@

In CUDA programming a kernel function is executed over a large set of threads, where the kernel function describes the work to be done in each thread. As depicted in Fig.\ \ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g. $32\times32$). Threads inside a block support low-cost synchronization if needed. From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of executing multiple arithmetic operations per clock cycle.
%32 or more concurrent multiplications or additions, or both, depending on the architecture. A group of blocks is scheduled to each SM until all blocks have been processed.

@@

\begin{figure}
\centerline{\includegraphics[width=3.5in]{gfx/gpu_layout_vertical_2.eps}}
\caption{Calculation of Capon adaptive weights mapped to GPU architecture. A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups. Note that several samples can be processed per thread-block if $L$ is small. The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
\label{fig:gpulayout}
\end{figure}

@@

\section{Parallel ES-Capon}\label{sec:meth}

@@

If the beamformer in (\ref{eq:z_mv}) is analyzed in combination with the matrix equation in (\ref{eq:b}) and the covariance estimation process in (\ref{eq:diag}), we see that the calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Fig.\ \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}.\\
\text{2) } &\text{Solve the linear system of equations as in (\ref{eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}.
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexities of these three steps are $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively. As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance in cardiac ultrasound imaging can only be achieved by exploring methods with reduced complexity.

@@

%\begin{figure}
%\includegraphics[width=3in]{gfx/gpu_layout_lowres.png}%
%\caption{}
%\label{fig:gpulayout}
%\end{figure}

\subsection{Calculation of Multiple Sample Covariance Matrices}\label{sec:calcR}

%The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ since $N_L$, the number of subarrays, is a function of $L$. This is actually the same as for inverting an $L \times L$ matrix. However little attention has been devoted to this step in the literature before. Note that for a typical array configuration we do more spacial smoothing than temporal smoothing. Hence, $N_K$ is usually small compared to $N_L$. The construction of $\mat{\hat{R}}$ is dominated by the number of elements $M$ ($N_L = M - L + 1$) for short subarrays, but since $L$ is selected in most cases to be between $M/2$ and $M/4$ a large $M$ implies a large $L$.

Calculation of one element of \R is from (\ref{eq:R}) found to be independent of calculation of the other elements. A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}. This will minimize global reads and writes needed per thread, but since we want to avoid block-to-block synchronization, the maximum supported subarray length is limited to 32 by this approach. Since most medical ultrasound arrays have 64 elements or more, an implementation not limited by $L_{max}=M/2=32$ is preferable.

In \cite{Asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ %and linear complexity with $N_K$ 
can be reduced by realizing that calculations in (\ref{eq:R}) overlap across subarrays. %Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
%\begin{align}
%\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
%&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
%&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
%\end{align}
%Along the time dimension we also find that
%\begin{align}
%\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
%&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
%\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
%Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$. However, since $N_K$ typically is small, and the GPU has a limited amount of shared memory available per thread, we focus on subarrays only. 
%In that case, 
The complexity of calculating $\mat{\hat{R}}$ is by this reduced to $O(N_K(L(N_L-2)+2L^2))$. For $L=M/2$ this reduces to $O(N_KM^2)$.
%The sliding-window approach will cause dependence among the iterative calculations. The level of task granularity is therefore increased. Calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore assigned to each diagonal in the upper triangle of $\mat{\hat{R}}$. For small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). 
Maximum supported $L$ is also increased to the maximum number of threads per block (currently 1024). %However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full GPU occupancy.

@@

%For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory. One row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements therefore need to reside in shared memory until the trace is accumulated. When that is done, scattered writes of the scaled diagonal elements to global memory has to be performed.

@@

\section{Benchmarks}\label{sec:bench}

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_64_80_540.png} \label{fig:benchUS64}}
\hfil
\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_96_120_540.png}%
\label{fig:benchUS96}}}
%\centerline{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_540.png}}
\caption{Benchmarks of GPU Capon beamforming in element space for a cardiac image covering a $70\degree$ wide $15.4$ cm deep sector. Execution time is plotted as a function of subarray length $L$. The figures show execution times for the three steps listed in Section \ref{sec:meth}: Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.}
\label{fig:bench}
\end{figure*}

>>>>

In the next section we present background information on Capon beamforming in both element-space and beamspace. We also give an introduction to the GPU computing model. In Section \ref{II_sec:meth} we give a summary of our parallel implementation of the element-space Capon beamformer, before the GPU implementation of the beamspace Capon beamformer is presented in Section \ref{II_sec:bs}. Benchmark results and resulting images for both implementations are presented in Section \ref{II_sec:bench} and \ref{II_sec:images}, together with a discussion of trading image quality for speed in Section \ref{II_sec:trade}. Finally we discuss the results in Section \ref{II_sec:dis}, and draw conclusions in Section \ref{II_sec:con}.

\section{Background}\label{II_background}

\subsection{Capon Beamforming}\label{II_sec:es-capon}

@@

The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS). Here, the signal recorded at element number $m \in [0,M-1]$ at time $n$ is delayed with an appropriate delay $\Delta_m[n]$ to focus and steer the ultrasound beam. This delayed data is here denoted $x_m[n]$. The beamformer output $z[n]$ is then the sum of all $M$ elements weighted with a factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n] = \vec{w}^H\vec{x}[n], \label{II_eq:z}
\end{align}
where $(\cdot)^H$ is the Hermitian transpose. The weight vector, or window, $\vec{w}$ is usually real in DAS beamforming, and is applied to trade resolution for a lower side lobe level. Although different windows can be used in different areas of the images, $\vec{w}$ is usually predefined.

As shown in Fig.\ \ref{II_fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.   The Capon beamformer produces weights that minimize the beamformer output power while maintaining unity gain in the focus direction. The result is that interference impinging from other directions is suppressed \cite{Synnevag2007}. As with coarse-fine beamformers \cite{Thomenius}, the complex weights cause a micro-steering which is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.

%Formally stated, the adaptive weights are found by solving the following minimization problem:
%\begin{align}
%&\min_{\vec{w}} E\{|z[n]|^2\} = \min_{\vec{w}}\vec{w}^H\mat{R}\vec{w}\label{II_eq:minz}\\
%&\text{subjected to } \vec{w}^H\vec{a} = 1,\label{II_eq:constraint}
%\end{align}
%where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$ is the spatial covariance matrix, and $\vec{a} = \left[ \begin{array}{cccc} e^{-jk_{\tilde{x}}\tilde{x}_0}& e^{-jk_{\tilde{x}}\tilde{x}_1}& ...& e^{-jk_{\tilde{x}}\tilde{x}_{M-1}}\end{array}\right] ^T$ is a far-field steering vector towards angle $\theta = \arcsin(\frac{\lambda k_{\tilde{x}}}{2\pi})$.
%The constraint in (\ref{II_eq:constraint}) is there to avoid the trivial solution $\vec{w} = 0$, and a distorted respons in the direction of $\vec{a}$. 
%The solution to the minimization problem in (\ref{II_eq:minz}) and (\ref{II_eq:constraint}) is

\begin{figure}
\centerline{\includegraphics[width=.7\linewidth]{gfx/beamforming_mv.eps}}
\caption{Capon beamforming. The impinging signal is demodulated, and aligned in phase by applying steering and focusing delays $\Delta_m[n]$. Adaptive weights, $\vec{w}[n]$, are then calculated based on in-phase data, and finally the output is formed by a Capon-weighted coherent sum.}
\label{II_fig:mvbf}
\end{figure}

The Capon beamformer, following the approach in \cite{Synnevag2009}, can be divided into three main steps: estimation of a sample covariance matrix, solving a system of linear equations, and calculation of the beamformer output.  The sample covariance matrix is estimated as 
%
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{II_eq:R}
\end{align}
%
where  $N_K = 2K + 1$ is the number of temporal samples included, $N_L = M-L+1$ is the number of subarrays, and $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l\text{th}$ subarray of length $L$. Note that the covariance matrix dimension is $L \times L$. The covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness:
%
\begin{align}\label{II_eq:diag}
\mat{\hat{R}}[n] = \mat{\breve{R}}[n] + \epsilon[n]\mat{I} = f(\vec{x}[n]).
\end{align}
%
The operator $f(\cdot)$ represents the process of estimating $\mat{\hat{R}}$ from the input $\vec{x}$.
A factor proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
%
\begin{align}\label{II_eq:diag_adapt}
\epsilon[n] &= d \times \frac{\text{trace}\{\mat{\breve{R}}[n]\}}{L},
\end{align}
%
has been applied in much of the recent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}. It has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.

Next, a linear system of equations has to be solved,
%
\begin{align}\label{II_eq:b}
\vec{b}[n] = \mat{\hat{R}}[n]^{-1}\vec{a} \in \mathbb{C}^L,
\end{align}
%
where the steering vector $\vec{a} = \vec{1}_L = [1_0, 1_1, ..., 1_{L-1}]^T$ since $\vec{x}$ is pre-delayed.

Finally the Capon weight vector is formed from (\ref{II_eq:b}) as
%
\begin{align}\label{II_eq:w}
\vec{w}[n] = \frac{\vec{b}[n]}{\vec{1}^H\vec{b}[n]} \in \mathbb{C}^L,
\end{align}
%
and the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights. This is formally stated as
%
\begin{align}
z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l[n] \label{II_eq:z_mv}\\
&= 1/N_L \, (\vec{w}[n] * \vec{1}_{N_L})^H\vec{x}[n] \label{II_eq:z_mv2},
\end{align}
%where (this matrix is wrong)
%\begin{align}
%\mat{C} = \left(
%\begin{matrix}
%\vec{1}_1 & \vec{1}_2 & \cdots & \vec{1}_{N_L} & \vec{0}_1 & \cdots & \vec{0}_{N_L-1} & \vec{0}_{N_L}\\
%\vec{0}_{N_L-1} & \vec{0}_{N_L-2} & \cdots & \vec{0}_{0} & \vec{1}_{N_L-1} & \cdots & \vec{1}_{2} & \vec{1}_{1}\\
%\end{matrix}
%\right)^T
%\end{align}
where $*$ is a non-truncated convolution across elements.
Note how we can either choose to sum all subarrays (\ref{II_eq:z_mv}), or accumulate a weighting per element (\ref{II_eq:z_mv2}).

In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{II_eq:R}) is typically averaged over a long sequence of samples\cite{Krima}. Since medical ultrasound imaging is an active, short-pulsed system, there is a limited amount of data from which $\mat{\hat{R}}$ can be estimated. The estimated matrix is therefore often found to be both inaccurate and poorly conditioned. Another challenge with active systems is the high correlation between signal and interference, resulting in signal cancellation \cite{Reddy1987}. Thus, the beamformer output can be lower than expected, even when there is signal present and the distortionless-response criterion is applied.

@@

%\begin{figure}
%\includegraphics[width=3in]{gfx/beamforming_mv_lowres.png}
%\caption{}
%\label{II_fig:mv}
%\end{figure}

@@

\subsection{Beamspace Processing}\label{II_bs-capon}
Nilsen and Hafizovic \cite{Nilsen2009} have proposed to apply beamspace processing to reduce the required computations of the Capon beamformer when applied to medical ultrasound imaging. Beamspace is an alternative representation of the per-element data, transformed into a fan of beams covering the sector illuminated by transmit beams. The transformation is formally stated as
\begin{align}
\vec{x}_\text{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \ \ \ \ \ \ \ \ \ \ \, \mat{B} \in \mathbb{C}^{N,M} \label{II_eq:butler} \\ 
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M} \ \ \ N \le M.
\end{align} 
The steering vectors in $\mat{B}$, the so-called Butler or normalized discrete fourier transform (DFT) matrix, determine each beam's direction in space. Beams where no interference is present can therefore be removed by simply removing rows from $\mat{B}$.

@@

The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{II_eq:w}) with the beamspace covariance matrix 
\begin{align}
\mat{\hat{R}}_\text{BS} = f(\vec{x}_\text{BS}) = f(\mat{B}\vec{x}), 
\end{align}
where $\mat{B} \in \mathbb{C}^{N_b,L}$, $N_b$ is the number of selected beams, and $f(\cdot)$ represent the covariance estimation process as defined in (\ref{II_eq:diag}).

In the remaining sections, the methods described in this section and Section \ref{II_sec:es-capon} will be referred to as beamspace Capon (BS-Capon) and element-space Capon (ES-Capon) respectively. The name Capon will refer to both methods.

@@

In CUDA programming a kernel function is executed over a large set of threads, where the kernel function describes the work to be done in each thread. As depicted in Fig.\ \ref{II_fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g. $32\times32$). Threads inside a block support low-cost synchronization if needed. From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of executing multiple arithmetic operations per clock cycle.
%32 or more concurrent multiplications or additions, or both, depending on the architecture. A group of blocks is scheduled to each SM until all blocks have been processed.

@@

\begin{figure}
\centerline{\includegraphics[width=3.5in]{gfx/gpu_layout_vertical_2.eps}}
\caption{Calculation of Capon adaptive weights mapped to GPU architecture. A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups. Note that several samples can be processed per thread-block if $L$ is small. The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
\label{II_fig:gpulayout}
\end{figure}

@@

\section{Parallel ES-Capon}\label{II_sec:meth}

@@

If the beamformer in (\ref{II_eq:z_mv}) is analyzed in combination with the matrix equation in (\ref{II_eq:b}) and the covariance estimation process in (\ref{II_eq:diag}), we see that the calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Fig.\ \ref{II_fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{II_eq:diag})}.\\
\text{2) } &\text{Solve the linear system of equations as in (\ref{II_eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{II_eq:z_mv})}.
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{II_eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexities of these three steps are $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively. As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance in cardiac ultrasound imaging can only be achieved by exploring methods with reduced complexity.

@@

%\begin{figure}
%\includegraphics[width=3in]{gfx/gpu_layout_lowres.png}%
%\caption{}
%\label{II_fig:gpulayout}
%\end{figure}

\subsection{Calculation of Multiple Sample Covariance Matrices}\label{II_sec:calcR}

%The straightforward sum in (\ref{II_eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ since $N_L$, the number of subarrays, is a function of $L$. This is actually the same as for inverting an $L \times L$ matrix. However little attention has been devoted to this step in the literature before. Note that for a typical array configuration we do more spacial smoothing than temporal smoothing. Hence, $N_K$ is usually small compared to $N_L$. The construction of $\mat{\hat{R}}$ is dominated by the number of elements $M$ ($N_L = M - L + 1$) for short subarrays, but since $L$ is selected in most cases to be between $M/2$ and $M/4$ a large $M$ implies a large $L$.

Calculation of one element of \R is from (\ref{II_eq:R}) found to be independent of calculation of the other elements. A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}. This will minimize global reads and writes needed per thread, but since we want to avoid block-to-block synchronization, the maximum supported subarray length is limited to 32 by this approach. Since most medical ultrasound arrays have 64 elements or more, an implementation not limited by $L_{max}=M/2=32$ is preferable.

In \cite{Asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ %and linear complexity with $N_K$ 
can be reduced by realizing that calculations in (\ref{II_eq:R}) overlap across subarrays. %Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
%\begin{align}
%\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
%&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
%&- x_{i-1}[n']x_{j-1}[n']^H). \label{II_eq:sliding}
%\end{align}
%Along the time dimension we also find that
%\begin{align}
%\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
%&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
%\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
%Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$. However, since $N_K$ typically is small, and the GPU has a limited amount of shared memory available per thread, we focus on subarrays only. 
%In that case, 
The complexity of calculating $\mat{\hat{R}}$ is by this reduced to $O(N_K(L(N_L-2)+2L^2))$. For $L=M/2$ this reduces to $O(N_KM^2)$.
%The sliding-window approach will cause dependence among the iterative calculations. The level of task granularity is therefore increased. Calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore assigned to each diagonal in the upper triangle of $\mat{\hat{R}}$. For small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{II_fig:gpulayout}). 
Maximum supported $L$ is also increased to the maximum number of threads per block (currently 1024). %However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full GPU occupancy.

@@

%For the adaptive diagonal loading factor in (\ref{II_eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory. One row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements therefore need to reside in shared memory until the trace is accumulated. When that is done, scattered writes of the scaled diagonal elements to global memory has to be performed.

@@

\section{Benchmarks}\label{II_sec:bench}

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_64_80_540.png} \label{II_fig:benchUS64}}
\hfil
\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_96_120_540.png}%
\label{II_fig:benchUS96}}}
%\centerline{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_540.png}}
\caption{Benchmarks of GPU Capon beamforming in element space for a cardiac image covering a $70\degree$ wide $15.4$ cm deep sector. Execution time is plotted as a function of subarray length $L$. The figures show execution times for the three steps listed in Section \ref{II_sec:meth}: Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.}
\label{II_fig:bench}
\end{figure*}
